Certified Knowledge Compilation \\ with Application to Formally Verified Model Counting

### 2024-01-07 doc/JAIR/lean-section.tex
add citations and fix references in Lean section

We set out to formally verify the system with two goals in mind:
first, to ensure that the CPOG framework is mathematically sound;
and second, to implement correct-by-construction proof checking
and ring evaluation (the ``Trusted Code'' components of Figure~\ref{fig:chain}).
These two goals are achieved with a single proof development
in the \lean{} programming language~\cite{demoura:cade:2021}.
Verification was greatly aided
by the Aesop~\cite{23limperg_aesop_white_box_best_first_proof_search_lean}
automated proof search tactic.
\lean{} is based on a logical foundation
in which expressions have a computational interpretation.
As in other proof assistants such as Isabelle and Coq,
functions defined in the formal system
can be compiled to machine code.
At the same time,
we can state and prove claims about them within the same system,
thereby verifying that our functions compute the intended results.
In this section,
we describe the functionality we implemented,
what we proved about it,
and the assumptions we made.

@@

On the mathematical side,
our specifications rely on a general theory of propositional logic
mirroring \ref{sect:logical:foundations}.
The type \lstinline{PropForm} describes the syntax of propositional formulas.
It is generic over the type of variables,
so we instantiate it with numeric variables as \lstinline{PropForm Var}.
Assignments of truth values are taken to be total functions \lstinline{PropAssignment Var := Var → Bool}.
Requiring totality is not a limitation:
instead of talking about two equal,
partial assignments to a subset $X' \subseteq X$ of variables,
we can more conveniently talk about two total assignments that agree on $X'$.
We write \lstinline{σ ⊨ φ} when \lstinline{σ : PropAssignment Var} satisfies \lstinline{φ : PropForm Var}.

@@

\paragraph{Proof checking.}
The goal of a CPOG proof is to construct a POG
that is equivalent to the input CNF $\inputformula$.
The database of active clauses,
the POG being constructed,
and its root literal,
are stored in a checker state structure {\tt PreState}.
The checker begins by parsing the input formula,
initializing the active clauses to $\theta \leftarrow \inputformula$,
and initializing the POG $P$ to an empty one.
It then processes every step of the CPOG proof,
either modifying its state by adding/deleting clauses in $\theta$
and adding nodes to $P$,
or throwing an exception if a step is incorrect.
Afterwards, it carries out the \textsc{final conditions} check of \ref{subsection:semantics}.

Throughout the process,
we maintain invariants needed to establish the final result.
These ensure that $P$ is partitioned
and that a successful final check entails the logical equivalence
of $\inputformula$ and $\phi_\noder$,
where $\noder$ is the final POG root (\ref{thm:lean:equiv}).
Formally, we define a type {\tt State}
consisting of those {\tt PreState}s
that satisfy all the invariants.
A {\tt State} is a structure
combining {\tt PreState} fields with additional ones
storing computationally irrelevant \emph{ghost state}
that asserts the invariants.
The fields of \lstinline{st : PreState} include
\lstinline{st.inputCnf} for $\inputformula$,
\lstinline{st.clauseDb} for $\theta$,
and \lstinline{st.pog} for $P$.
We write \lstinline{st.pogDefsForm} for the clausal POG definitions formula $\bigwedge_{\nodeu\in P}\theta_u$,
and \lstinline{st.allVars} for all variables (original and extension) added so far.
For any $\nodeu\in P$,
\lstinline{st.pog.toPropForm u} computes $\phi_\nodeu$.

@@

invariants.equivInput : equivalentOver st.inputCnf.vars st.inputCnf st.clauseDb
\end{lstlisting}
Finally, for every node $\nodeu\in P$ with corresponding literal $u$ we ensure that $\phi_\nodeu$ is partitioned (\ref{def:partitioned-operation-formula}) and relate $\phi_\nodeu$ to its clausal encoding $\theta_u \doteq u \wedge \bigwedge_{\nodev\in P}\theta_v$:
\begin{lstlisting}
def partitioned : PropForm Var → Prop
  | tr | fls | var _ => True
  | neg φ    => φ.partitioned
  | disj φ ψ => φ.partitioned ∧ ψ.partitioned ∧ ∀ τ, ¬(τ ⊨ φ ∧ τ ⊨ ψ)
  | conj φ ψ => φ.partitioned ∧ ψ.partitioned ∧ φ.vars ∩ ψ.vars = ∅

@@

\paragraph{Ring evaluation.} We formalized the central quantity (\ref{eqn:rep}) in the ring evaluation problem
(\ref{def:ring_evaluation}) in a commutative ring \lstinline{R} as follows:
\begin{lstlisting}
def weightSum {R : Type} [CommRing R]
    (weight : Var → R) (φ : PropForm Var) (s : Finset Var) : R :=
  ∑ τ in models φ s, ∏ x in s, if τ x then weight x else 1 - weight x
\end{lstlisting}
The rules for efficient ring evaluation of partitioned formulas are expressed as:
\begin{lstlisting}
def ringEval (weight : Var → R) : PropForm Var → R
  | tr       => 1
  | fls      => 0
  | var x    => weight x
  | neg φ    => 1 - ringEval weight φ
  | disj φ ψ => ringEval weight φ + ringEval weight ψ
  | conj φ ψ => ringEval weight φ * ringEval weight ψ
\end{lstlisting}
\ref{prop:ring:eval} is then formalized as follows:
\begin{lstlisting}
theorem ringEval_eq_weightSum (weight : Var → R) {φ : PropForm Var} :
    partitioned φ → ringEval weight φ = weightSum weight φ (vars φ)
\end{lstlisting}
To efficiently compute the ring evaluation of a formula represented by a POG node, we implemented
\lstinline{Pog.ringEval} and then proved that it matches the specification above:
\begin{lstlisting}
theorem ringEval_eq_ringEval (pog : Pog) (weight : Var → R) (x : Var) :
  pog.ringEval weight x = (pog.toPropForm x).ringEval weight
\end{lstlisting}
Applying this to the output of our verified CPOG proof checker, which we know to be partitioned
and equivalent to the input formula $\inputformula$, we obtain a proof that our toolchain computes
the correct ring evaluation of $\inputformula$.

>>>>

We set out to formally verify the system with two goals in mind:
first, to ensure that the CPOG framework is mathematically sound;
and second, to implement correct-by-construction proof checking
and ring evaluation (the ``Trusted Code'' components of Figure~\ref{fig:chain}).
These two goals are achieved with a single proof development
in the \lean{} programming language~\cite{demoura:cade:2021}.
Verification was greatly aided
by the Aesop~\cite{23limperg_aesop_white_box_best_first_proof_search_lean}
automated proof search tactic.
\lean{} is based on a logical foundation
in which expressions have a computational interpretation.
As in other proof assistants such as Isabelle~\cite{nipkow:et:al:02} and Coq~\cite{coq},
functions defined in the formal system
can be compiled to machine code.
At the same time,
we can state and prove claims about them within the same system,
thereby verifying that our functions compute the intended results.
In this section,
we describe the functionality we implemented,
what we proved about it,
and the assumptions we made.

@@

On the mathematical side,
our specifications rely on a general theory of propositional logic
mirroring Section~\ref{sect:logical:foundations}.
The type \lstinline{PropForm} describes the syntax of propositional formulas.
It is generic over the type of variables,
so we instantiate it with numeric variables as \lstinline{PropForm Var}.
Assignments of truth values are taken to be total functions \lstinline{PropAssignment Var := Var → Bool}.
Requiring totality is not a limitation:
instead of talking about two equal,
partial assignments to a subset $X' \subseteq X$ of variables,
we can more conveniently talk about two total assignments that agree on $X'$.
We write \lstinline{σ ⊨ φ} when \lstinline{σ : PropAssignment Var} satisfies \lstinline{φ : PropForm Var}.

@@

\paragraph{Proof checking.}
The goal of a CPOG proof is to construct a POG
that is equivalent to the input CNF $\inputformula$.
The database of active clauses,
the POG being constructed,
and its root literal,
are stored in a checker state structure {\tt PreState}.
The checker begins by parsing the input formula,
initializing the active clauses to $\theta \leftarrow \inputformula$,
and initializing the POG $P$ to an empty one.
It then processes every step of the CPOG proof,
either modifying its state by adding/deleting clauses in $\theta$
and adding nodes to $P$,
or throwing an exception if a step is incorrect.
Afterwards, it carries out the \textsc{final conditions} check of Section~\ref{subsection:semantics}.

Throughout the process,
we maintain invariants needed to establish the final result.
These ensure that $P$ is partitioned
and that a successful final check entails the logical equivalence
of $\inputformula$ and $\phi_\noder$,
where $\noder$ is the final POG root (Theorem~\ref{thm:lean:equiv}).
Formally, we define a type {\tt State}
consisting of those {\tt PreState}s
that satisfy all the invariants.
A {\tt State} is a structure
combining {\tt PreState} fields with additional ones
storing computationally irrelevant \emph{ghost state}
that asserts the invariants.
The fields of \lstinline{st : PreState} include
\lstinline{st.inputCnf} for $\inputformula$,
\lstinline{st.clauseDb} for $\theta$,
and \lstinline{st.pog} for $P$.
We write \lstinline{st.pogDefsForm} for the clausal POG definitions formula $\bigwedge_{\nodeu\in P}\theta_u$,
and \lstinline{st.allVars} for all variables (original and extension) added so far.
For any $\nodeu\in P$,
\lstinline{st.pog.toPropForm u} computes $\phi_\nodeu$.

@@

invariants.equivInput : equivalentOver st.inputCnf.vars st.inputCnf st.clauseDb
\end{lstlisting}
Finally, for every node $\nodeu\in P$ with corresponding literal $u$ we ensure that $\phi_\nodeu$ is partitioned (Definition~\ref{def:partitioned-operation-formula}) and relate $\phi_\nodeu$ to its clausal encoding $\theta_u \doteq u \wedge \bigwedge_{\nodev\in P}\theta_v$:
\begin{lstlisting}
def partitioned : PropForm Var → Prop
  | tr | fls | var _ => True
  | neg φ    => φ.partitioned
  | disj φ ψ => φ.partitioned ∧ ψ.partitioned ∧ ∀ τ, ¬(τ ⊨ φ ∧ τ ⊨ ψ)
  | conj φ ψ => φ.partitioned ∧ ψ.partitioned ∧ φ.vars ∩ ψ.vars = ∅

@@

\paragraph{Ring evaluation.} We formalized the central quantity (\ref{eqn:rep}) in the ring evaluation problem
(Definition \ref{def:ring_evaluation}) in a commutative ring \lstinline{R} as follows:
\begin{lstlisting}
def weightSum {R : Type} [CommRing R]
    (weight : Var → R) (φ : PropForm Var) (s : Finset Var) : R :=
  ∑ τ in models φ s, ∏ x in s, if τ x then weight x else 1 - weight x
\end{lstlisting}
The rules for efficient ring evaluation of partitioned formulas are expressed as:
\begin{lstlisting}
def ringEval (weight : Var → R) : PropForm Var → R
  | tr       => 1
  | fls      => 0
  | var x    => weight x
  | neg φ    => 1 - ringEval weight φ
  | disj φ ψ => ringEval weight φ + ringEval weight ψ
  | conj φ ψ => ringEval weight φ * ringEval weight ψ
\end{lstlisting}
Proposition~\ref{prop:ring:eval} is then formalized as follows:
\begin{lstlisting}
theorem ringEval_eq_weightSum (weight : Var → R) {φ : PropForm Var} :
    partitioned φ → ringEval weight φ = weightSum weight φ (vars φ)
\end{lstlisting}
To efficiently compute the ring evaluation of a formula represented by a POG node, we implemented
\lstinline{Pog.ringEval} and then proved that it matches the specification above:
\begin{lstlisting}
theorem ringEval_eq_ringEval (pog : Pog) (weight : Var → R) (x : Var) :
  pog.ringEval weight x = (pog.toPropForm x).ringEval weight
\end{lstlisting}
Applying this to the output of our verified CPOG proof checker, which we know to be partitioned
and equivalent to the input formula $\inputformula$, we obtain a proof that our toolchain computes
the correct ring evaluation of $\inputformula$.
